---
title: "CMH Issue Demo"
author: "Mike Stackhouse"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Backstory

Within the original [CDISC Pilot Project](https://bitbucket.cdisc.org/projects/CED/repos/sdtm-adam-pilot-project/browse), the example CSR included the use of the CMH test. For portions of the analysis, the statistical analysis plan specifically calls out the following:

> Treatments will be compared for overall differences by Cochran-Mantel-Haentzel (CMH) test referred to in SAS(R) as “row mean scores differ,” 2 degrees of freedom. 

While base R has the [`mantelhaen.test`](https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/mantelhaen.test) function, unfortunately the alternative hypothesis specified in the CDISC Pilot Project is not available.

There are several other packages that offer implementations of the CMH test, but the only one (at least that I have found) that offers the "row mean scores differ" alternative hypothesis is the package [`vcdExtra`](https://github.com/friendly/vcdExtra) through the function [`vcdExtra::CMHtest`](https://www.rdocumentation.org/packages/vcdExtra/versions/0.7-1/topics/CMHtest).

Unfortunately, a problem arises with the CDISC pilot data, as can be seen in [this GitHub issue](https://github.com/friendly/vcdExtra/issues/3):

> For large sparse table (many strata) CMHTest will occasionally throw a Error in solve.default(AVA)

This error is encountered when trying to replicate outputs in the CDISC pilot project such as [Table 14-3.13](https://github.com/atorus-research/CDISC_pilot_replication/blob/master/outputs/14-3.13.rtf). The thread goes on to recommend that a use of `solve()` is replaced with `MASS::ginv`.

This update does in fact work, and I've implemented in the fork of the `vcdExtra` package below. This can be demonstrated in the following example

_Note: For further examples in this document to work you will need a forked version of the `vcdExtra` package installed. You can install as follows:_

```{r VCDExtra install, eval=FALSE}
Sys.setenv("R_REMOTES_NO_ERRORS_FROM_WARNINGS" = "true")
devtools::install_github("mstackhouse/vcdExtra")
```

## Data Preparation

The data here are taken from the the [PHUSE Test Data Factory](https://advance.phuse.global/display/WEL/Test+Dataset+Factory) replication of the [CDISC Pilot Project](https://bitbucket.cdisc.org/projects/CED/repos/sdtm-adam-pilot-project/browse). 

```{r dataprep}
cbic <- haven::read_xpt("https://github.com/phuse-org/TestDataFactory/raw/main/Updated/TDF_ADaM/adcibc.xpt") %>%
  filter(EFFFL == 'Y' & ITTFL == 'Y', AVISITN %in% c(8, 16, 24) & ANL01FL=='Y')
```

## Row Mean Scores Differ Example

With the subset data included, we can create a function for the CMH test using the forked vcdExtra package. 

```{r cmh_function}
cmh_p <- function(.data, formula, alternate=c('rmeans', 'cmeans', 'general', 'cor')) {

  # Pull out the hypoth
  alternate <- match.arg(alternate, several.ok=FALSE)

  # Run the test
  res <- vcdExtra::CMHtest(formula, data=.data, overall=TRUE)$ALL

  pvalue <- unlist(res$table[alternate, 'Prob'])
  pvalue
}

```

And we can recreate the results from the CDISC pilot.

```{r}
cbic %>% 
  group_by(AVISITN) %>% 
  group_map(~ cmh_p(.x, formula=AVAL ~ TRTP | SITEGR1)) %>%
  unlist()
```
