---
title: "MMRM - SAS versus R"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
This document focuses on a comparison of results generated using a Mixed Model for Repeated Measures in SAS and R. Data for the comparison was the lab ADaM dataset adlbh.xpt from the Phuse Pilot Study. Results were generated for each lab parameter and time point in the dataset using three different covariance structures, i.e. unstructured, compound symmetry and autoregressive of first order (AR(1)).

## SAS code
In SAS the following code was used (assessments at avisitn=0 should also be removed from the response variable):
<pre><code>
proc mixed data=adlbh;
  where base ne . and avisitn not in (., 99);
  class usubjid trtpn(ref="0") avisitn;
  by paramcd param;
  model chg=base trtpn avisitn  trtpn*avisitn / solution cl alpha=0.05 ddfm=KR;
  repeated avisitn/subject=usubjid type=&covar;
  lsmeans trtpn * avisitn / diff cl slice=avisitn;
  lsmeans trtpn / diff cl;
run;
</code></pre>

where the macro variable covar could be UN, CS or AR(1). The results were stored in .csv files that were post-processed in R and compared with the results from R.

## R code using the gls function
The code below implements the MMRM using the gls function from the nlme package. Afterwards the emmeans package is used to get LSmeans estimates.

<pre><code>
### Unstructured
mod1<-tryCatch(gls(model = CHG ~ TRTP + AVISITN + TRTP:AVISITN + AVISITN + BASE,
                   data = adlbh_1,
                   correlation = corSymm(form = ~1|SUBJID),
                   weights = varIdent(form = ~1|AVISITN),
                   control = glsControl(opt = "optim"),
                   method = "REML",
                   na.action = "na.omit"),
                   error = function(x) return(NULL))

### Compound symmetry
mod1<-tryCatch(gls(model = CHG ~ TRTP + AVISITN + TRTP:AVISITN + AVISITN + BASE,
                   data = adlbh_1,
                   correlation = corCompSymm(form= ~ 1 | SUBJID),
                   weights = varIdent(form = ~1|AVISITN),
                   control = glsControl(opt = "optim"),
                   method = "REML",
                   na.action = "na.omit"),
                   error = function(x) return(NULL))

### AR(1)
mod1<-tryCatch(gls(model = CHG ~ TRTP + AVISITN + TRTP:AVISITN + AVISITN + BASE,
                   data = adlbh_1,
                   correlation = corCAR1(form= ~ 1 | SUBJID),
                   weights = varIdent(form = ~1|AVISITN),
                   control = glsControl(opt = "optim"),
                   method = "REML",
                   na.action = "na.omit"),
                   error = function(x) return(NULL))

### Get LSmeans
emmeans(mod1, ~TRTP|AVISITN, mode="df.error")
emmeans::contrast(mod1, method="pairwise", adjust=NULL)

</pre></code>

## Comparison between SAS and R using the gls function
We then graphically compared the results from SAS with the results from R using the gls function. 

```{r, echo=FALSE}
path<-"D:/working group/"
path1<-"D:/working group/results_first/"
from_R<-read.csv(paste0(path,"pair_diff.csv"))
from_SAS<-read.csv(paste0(path1,"LSMeansDifferences_ANCOVAwithRepeatedMeasures_KR_ADLBH_GithubData.csv"))

### Data manipulation of the SAS output 
from_SAS$contrast<-paste(from_SAS$reftrtp,"-",from_SAS$TRTP)
from_SAS$contrast[from_SAS$contrast=="Xanomeline High Dose - Xanomeline Low Dose"]<-"Xanomeline Low Dose - Xanomeline High Dose"
from_SAS$VISIT<-toupper(from_SAS$AVISIT)
from_SAS$corr<-from_SAS$covariatestructure
from_SAS$corr[from_SAS$corr=="First Order Autoregressive"]<-"corCAR1"
from_SAS$corr[from_SAS$corr=="Compound Symmetry"]<-"corCompSymm"
from_SAS$corr[from_SAS$corr=="Unstructured"]<-"corSymm"
from_SAS_1<-subset(from_SAS,corr %in% c("corCAR1","corCompSymm","corSymm"),select=c("PARAMCD","corr","contrast","VISIT","lsmeans_difference","lowerci95","upperci95"))

### Data manipulation of the R output 
from_R$VISIT<-paste("WEEK",from_R$AVISITN)
from_R$upper<-from_R$upper.CL
from_R$lower<-from_R$lower.CL
for(i in 1:dim(from_R)[1]){
  if(grepl("Placebo",from_R$contrast[i])){
    from_R$estimate[i]<-(-1)*from_R$estimate[i]
    from_R$lower[i]<-(-1)*from_R$upper.CL[i]
    from_R$upper[i]<-(-1)*from_R$lower.CL[i]
  }
}

from_R_1<-subset(from_R,select=c("PARAMCD","corr","contrast","VISIT","estimate","lower","upper"))

### Merge with output from R (by paramcd, visit, correlation structure and contrast)
temp01<-merge(from_SAS_1,from_R_1,by=c("PARAMCD","corr","contrast","VISIT"))
```

We then plotted the estimates of the treatment effect against each other  
```{r, echo=FALSE}
par(cex=1.3)
plot(temp01$lsmeans_difference,temp01$estimate,xlab="From SAS",ylab="From R",main="Estimates of treatment effect")
lines(c(-10,15),c(-10,15),lty=2)
points_to_plot<-subset(temp01,corr=="corCAR1")
points(points_to_plot$lsmeans_difference,points_to_plot$estimate,pch=19,col="red")
points_to_plot<-subset(temp01,corr=="corCompSymm")
points(points_to_plot$lsmeans_difference,points_to_plot$estimate,pch=19,col="green")
points_to_plot<-subset(temp01,corr=="corSymm")
points(points_to_plot$lsmeans_difference,points_to_plot$estimate,pch=19,col="blue")
legend("bottomright",c("AR(1)","Compound Symmetry","Unstructured"),col=c("red","green","blue"),pch=19)
```

The widths of the confidence intervals produced by SAS and R were also plotted aginst each other

```{r, echo=FALSE}
### Find width of 95% CI and compute summary statistics
temp01$width_CI_SAS<-temp01$upperci95-temp01$lowerci95
temp01$width_CI_R<-temp01$upper-temp01$lower

### Plot width of 95% CI against each other
par(cex=1.3)
plot(c(0,50),c(0,50),type="l",lty=2,xlab="From SAS",ylab="From R",main="Width of 95% CI")
points(temp01$width_CI_SAS,temp01$width_CI_R)
points_to_plot<-subset(temp01,corr=="corCAR1")
points(points_to_plot$width_CI_SAS,points_to_plot$width_CI_R,pch=19,col="red")
points_to_plot<-subset(temp01,corr=="corCompSymm")
points(points_to_plot$width_CI_SAS,points_to_plot$width_CI_R,pch=19,col="green")
points_to_plot<-subset(temp01,corr=="corSymm")
points(points_to_plot$width_CI_SAS,points_to_plot$width_CI_R,pch=19,col="blue")
legend("bottomright",c("AR(1)","Compound Symmetry","Unstructured"),col=c("red","green","blue"),pch=19)
```

## R code using the lmer function
The repeated measures model with an unstructured covariance matrix can also be formulated as a random effect model and implemented using the lmer function from the lme4 package. This is shown below

<pre><code>
mod1<-lmer(CHG ~ TRTP + AVISITN + TRTP:AVISITN + BASE + (0 + AVISITN | SUBJID),data=adlbh_1, control=lmerControl(check.nobs.vs.nRE="ignore"), na.action=na.omit)
</pre></code>

## Comparison between SAS and R using the lmer function
We then graphically compared the results from SAS with the results from R using the lmer function. Again, estimates of treatment effect and width of confidence intervals were plotted against each other.

```{r, echo=FALSE}
### Compare with SAS
final_1<-read.csv(paste0(path,"roche_diff.csv"))
from_SAS_2<-subset(from_SAS_1,corr=="corSymm")
temp02<-merge(from_SAS_2,final_1,by=c("PARAMCD","contrast","VISIT"))

### Plot estimates against each other
par(cex=1.3)
plot(temp02$lsmeans_difference,temp02$estimate,xlab="From SAS",ylab="From R (Roche approach)",main="Estimates of treatment effect")
lines(c(-10,15),c(-10,15),lty=2)

### Find width of 95% CI and compute summary statistics
temp02$width_CI_SAS<-temp02$upperci95-temp02$lowerci95
temp02$width_CI_R<-temp02$upper-temp02$lower

### Plot width of 95% CI against each other
par(cex=1.3)
plot(c(0,40),c(0,40),type="l",lty=2,xlab="From SAS",ylab="From R (Roche approach)",main="Width of 95% CI")
points(temp02$width_CI_SAS,temp02$width_CI_R)
```

## Future work
+ Run SAS code by also removing assessments at avisitn=0 from the response variable, and using trtp (or trtpn) and avisit (or avisitn)
+ Investigating the differences
+ Implement lmer equivalent to MMRM with compound symmetry
+ Comparisons for other models, i.e. only random, random and repeated, no repeated

